---
title: "sql-workshop"
author: "Charles Lang"
output: html_document
---

Before you follow the directions below, please take a screenshot of your AWS console showing the running database and upload it to your repo.

## Connect to AWS MySQL Database
```{r}
#install.packages("DBI", "RMySQL")

library(DBI)
library(RMySQL)

db_user <- 'admin'
db_password <- 'testsql!'
db_name <- 'oudb'
db_host <- 'database-1.cqyfynigtzsr.us-east-2.rds.amazonaws.com'
db_port <- 3306

mydb <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)

summary(mydb)
```

## Load OU Data
```{r}
#Student demographic data
studentInfo <- read.csv("studentInfo.csv", header = TRUE)
#Student assessment data
studentAssessment <- read.csv("studentAssessment.csv", header = TRUE)
#Course data
courses <- read.csv("courses.csv", header = TRUE)
studentRegistration <- read.csv("studentRegistration.csv", header = TRUE)
```

## Write data to the DB using the DBI package
```{r}
#List the tables in the DB - should be zero
dbListTables(mydb)

#Write a new table to the DB
dbWriteTable(mydb, "studentInfo", studentInfo)
dbWriteTable(mydb, "studentAssessment", studentAssessment)
dbWriteTable(mydb, "courses", courses)
dbWriteTable(mydb, "studentRegistration", studentRegistration)

#List tables to see that table was added
dbListTables(mydb)

#Read a particular table
dbReadTable(mydb, 'studentInfo')

#EXERCISE 1
#Make two toy data sets with at least three variables and at least 30 rows each in them. Have a mix of numeric and character variables. Transfer these dataframes to your SQL database using the DBI commands. Name the tables whatever you like.

#Generate toy dataset 1: List of top 30 countries by GDP (PPP) per capita in 2020 according to the International Money Fund.

econ <- data.frame(country = c("Luxembourg", "Singapore", "Qatar", "Ireland", "Switzerland",
                               "Norway", "United States", "Brunei", "United Arab Emirates",
                               "Denmark", "Netherlands", "San Marino", "Austria", "Iceland",
                               "Germany", "Sweden", "Australia", "Belgium", "Finland",
                               "Bahrain", "Canada", "Saudi Arabia", "France", "South Korea",
                               "United Kingdom", "Malta", "Kuwait", "Japan", "New Zealand",
                               "Czech Republic"),
                   rank = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,
                            25,26,27,28,29,30),
                   gdp_pc =c(112875, 95603, 91897, 89383, 68340, 64856, 63051, 61816, 58466,
                             57781, 57101, 56690, 55406, 54482, 53571, 52477, 50845, 50114,
                             49334, 49057, 47569, 46273, 45454, 44292, 44288, 43087, 41735,
                             41637, 41072, 40293))

#Generate toy dataset 2: List of top 30 NYC charter schools by mean scale score on the 2018-19 grades 3-8 English language arts (ELA) assessment.

elameanscore <- data.frame(borough = c("Manhattan", "Manhattan", "Manhattan", "Manhattan",
                                       "Manhattan", "Bronx", "Brooklyn", "Manhattan",
                                       "Manhattan", "Bronx", "Brooklyn", "Manhattan",
                                       "Manhattan", "Bronx", "Brooklyn", "Manhattan",
                                       "Manhattan", "Manhattan", "Bronx", "Bronx",    
                                       "Brooklyn", "Brooklyn", "Brooklyn", "Brooklyn",  
                                       "Brooklyn", "Brooklyn", "Brooklyn", "Queens",  
                                       "Manhattan", "Manhattan"),
                           school = c("Success Academy Charter School-Upper West", "Success
                                      Academy Charter School-Union Square", "Success Academy
                                      Charter School-Union Square", "Success Academy Charter
                                      School-Upper West", "Success Academy Charter School -
                                      Washington Heights", "Success Academy Charter   
                                      School-Bronx 2", "Success Academy Charter School -    
                                      Bensonhurst", "Success Academy Charter School-Upper   
                                      West", "Success Academy Charter School-Harlem 5",  
                                      "Success Academy Charter School-Bronx 2", "Succes
                                      Academy Charter School-Cobble Hill", "Success Academy   
                                      Charter School-Upper West", "Success Academy Charter
                                      School-Harlem 3", "South Bronx Classical Charter School
                                      III", "Success Academy Charter School-Crown Heights",
                                      "Success Academy Charter School-Union Square", "Success
                                      Academy, Charter School-Union Square", "Success Academy
                                      Charter School-Upper West", "Family Life Academy Charter
                                      School II", "South Bronx Classical Charter School III",
                                      "Success Academy Charter School - Bed Stuy 1", "Success
                                      Academy Charter School-Bed Stuy 2", "Success Academy    
                                      Charter School-Williamsburg", "Bedford Stuyvesant New 
                                      Beginnings Charter School", "Success Academy Charter 
                                      School-Prospect Heights", "Success Academy Charter
                                      School-Prospect Heights", "Success Academy Charter
                                      School-Flatbush", "Success Academy Charter School - 
                                      Rosedale", "Success Academy Charter School-Hell's
                                      Kitchen", "Sisulu-Walker Charter School Of Harlem"),
                           score = c(626, 625, 624, 624, 624, 624, 624, 623, 623, 623, 623,
                                     622, 622, 622, 622, 621, 621, 621, 621, 621, 621, 621,
                                     621, 621, 621, 621, 621, 621, 620, 620),
                           grade = c(8,7,3,3,3,8,3,4,8,7,3,5,7,3,3,5,6,7,3,4,6,
                                     3,3,4,3,7,3,3,3,4))

#Generate toy dataset 3 for matching purposes: List of top 30 NYC charter schools by mean scale score on the 2018-19 grades 3-8 math assessment.

mathmeanscore <- data.frame(borough = c("Manhattan", "Manhattan", "Manhattan", "Manhattan",
                                       "Brooklyn", "Bronx", "Brooklyn", "Manhattan",
                                       "Manhattan", "Manhattan", "Manhattan", "Manhattan",
                                       "Manhattan", "Brooklyn", "Manhattan", "Manhattan",
                                       "Bronx", "Brooklyn", "Brooklyn", "Manhattan",    
                                       "Manhattan", "Manhattan", "Brooklyn", "Brooklyn",  
                                       "Brooklyn", "Manhattan", "Manhattan", "Manhattan",  
                                       "Manhattan", "Manhattan"),
                           school = c("Success Academy Charter School-Union Square","Success
                                      Academy Charter School-Union Square","Success Academy
                                      Charter School-Harlem 4","Success Academy Charter
                                      School-Upper West","Success Academy Charter
                                      School-Prospect Heights","Success Academy Charter
                                      School-Bronx 2","Achievement First-Bushwick Charter
                                      School","Success Academy Charter School-Harlem
                                      1","Success Academy Charter School-Harlem 3","Success
                                      Academy Charter School-Upper West","Success Academy
                                      Charter School-Upper West","Success Academy Charter
                                      School-Harlem 3","Success Academy Charter School-Harlem
                                      5","Success Academy Charter School -
                                      Bensonhurst","Success Academy Charter School-Union
                                      Square","Success Academy Charter School-Hell's
                                      Kitchen","Success Academy Charter School-Bronx
                                      2","Success Academy Charter School-Crown Heights",
                                      "Success Academy Charter School - Bensonhurst","Success
                                      Academy Charter School-Harlem 3","Success Academy
                                      Charter School-Harlem 2","Success Academy Charter
                                      School-Harlem 5","Success Academy Charter
                                      School-Williamsburg","Success Academy Charter
                                      School-Crown Heights","Success Academy Cs - Bergen
                                      Beach","Success Academy Charter School-Union
                                      Square","Success Academy Charter School-Union
                                      Square","Success Academy Charter School-Upper
                                      West","Success Academy Charter School - Washington
                                      Heights","Success Academy Charter School - Washington
                                      Heights"),
                           score = c(639, 636, 636, 635, 635, 634, 634, 633, 633, 632, 632,  
                                     632, 632, 632, 631, 631, 631, 631, 631, 630, 630, 630,
                                     630, 630, 630, 629, 629, 629, 629, 629),
                           grade = c(7,5,4,4,7,5,8,5,5,3,5,7,5,3,4,4,7,4,4,3,7,7,4,6,4,3,
                                     6,7,3,4))

#Add tables to the DB
dbWriteTable(mydb, "econ", econ)
dbWriteTable(mydb, "elameanscore", elameanscore, overwrite = T)
dbWriteTable(mydb, "mathmeanscore", mathmeanscore)

#List tables to see that tables were added
dbListTables(mydb)
```

## Getting into SQL - READING
```{r}
#Query a portion of the database (always returns dataframe)
dbGetQuery(mydb, "SELECT * FROM studentInfo LIMIT 10;")

dbGetQuery(mydb, "SELECT * FROM studentInfo ORDER BY id_student LIMIT 10;")

dbGetQuery(mydb, "SELECT id_student, gender FROM studentInfo ORDER BY id_student DESC LIMIT 10;") #Order listed will be reflected in order in table

dbGetQuery(mydb, "SELECT id_student AS 'Student ID', gender FROM studentInfo LIMIT 10;") #SQL Standard says quotes for literal strings and double quotes for everything else but that conflicts with R

#Count the number of rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#Using a WHERE statement on all columns
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50;")

#Using a WHERE statement on a single column (will not include missing data)
dbGetQuery(mydb, "SELECT COUNT(score) FROM studentAssessment WHERE score > 50;")

#Using an AND statement
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50 AND id_assessment = '1752';")

#EXERCISE 2
#Read toy data table 1, where gdp_pc is renamed and output is ordered in descending order and limited to the first 20 rows.

dbGetQuery(mydb, "SELECT gdp_pc AS 'GDP Per Capita', country FROM econ ORDER BY country DESC LIMIT 20;") 

#Read toy data table 2 to segment schools that are located in Manhattan.

dbGetQuery(mydb, "SELECT school, borough FROM elameanscore WHERE borough = 'Manhattan';")  

```

## Getting into SQL - UPDATING
```{r}
#Count rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#Add a row
dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted, is_banked, score) VALUES ('00001', '1', '20', '0', '50');")

#Count rows again
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Add a row with missing values
dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted) VALUES ('00001', '1', '20');")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Update a row
dbGetQuery(mydb, "UPDATE studentAssessment SET score = '20' WHERE id_student = 1;")

dbGetQuery(mydb, "SELECT id_student, score FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Update a row with NULL
dbGetQuery(mydb, "UPDATE studentAssessment SET score = 'NULL' WHERE id_student = 6516;")

#Delete a row (destructive)
dbGetQuery(mydb, "DELETE FROM studentAssessment WHERE id_student = 1;")

dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#EXERCISE 3
#Insert a new row in one of your toy data tables leaving one variable empty. Change one value in your other table. Display your new tables. Delete the row you edited and the row you inserted. 

#Add a row with missing values
dbGetQuery(mydb, "INSERT INTO econ (country, gdp_pc) VALUES ('Argentina', '382760');")
dbGetQuery(mydb, "SELECT * FROM econ ORDER BY country LIMIT 10;")
```

## Add/Deleting Table
```{r}
#Creating a new table in SQL
dbGetQuery(mydb,"CREATE TABLE test (
  score INTEGER, 
  student TEXT
  );")

dbListTables(mydb)

#Inserting data into the table
dbGetQuery(mydb, "INSERT INTO test VALUES ( 10, 'Amy' );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 11, 'Jen' );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 9, 'Frank' );")

dbGetQuery(mydb, "SELECT * FROM test;")

#Inserting a NULL row
dbGetQuery(mydb, "INSERT INTO test DEFAULT VALUES;") #Will not work use instead:

dbGetQuery(mydb,"INSERT INTO test (score, student) SELECT score, id_student FROM studentAssessment;")

#Delete a table
dbGetQuery(mydb, "DROP TABLE test;")

dbGetQuery(mydb, "SELECT * FROM test;") #This should produce an error since your table no longer exists

#Delete a table if it exists
dbGetQuery(mydb, "DROP TABLE IF EXISTS test;") #No error since it is only if it exists

#EXERCISE 4
#Create a table copy of first toy dataset using SQL commands. 
dbGetQuery(mydb,"CREATE TABLE elameanscore_test (
  borough TEXT, 
  school TEXT,
  score INTEGER
  );")

dbListTables(mydb)

#Inserting data into the table
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',626 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',625 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School - Washington Heights',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Bronx', 'Success Academy Charter School-Bronx 2',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School - Bensonhurst',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Harlem 5',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Bronx', 'Success Academy Charter School-Bronx 2',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Cobble Hill',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Harlem 3',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Bronx', 'South Bronx Classical Charter School III',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Crown Heights',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Bronx', 'Family Life Academy Charter School II',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Bronx', 'South Bronx Classical Charter School III',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School - Bed Stuy 1',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Bed Stuy 2',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Williamsburg',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Bedford Stuyvesant New Beginnings Charter School',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Prospect Heights',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Prospect Heights', 621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Brooklyn', 'Success Academy Charter School-Flatbush',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Queens', 'Success Academy Charter School - Rosedale',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Success Academy Charter School-Hells Kitchen',620 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test VALUES ( 'Manhattan', 'Sisulu-Walker Charter School of Harlem',620 );")

#Display toy dataset 1.
dbGetQuery(mydb, "SELECT * FROM elameanscore_test;")

#Delete the original table.

dbGetQuery(mydb, "DROP TABLE elameanscore_test;")
dbGetQuery(mydb, "SELECT * FROM elameanscore_test;")

```

# NULL Value
```{r}
#NULL is a state (similar to R), represents the lack of a value. But is not compatible with R backend so this code doesn't work as part of dbGetQuery()

#This doesn't work because NULL is not a value
SELECT * FROM test WHERE score = NULL;

#Instead use
SELECT * FROM test WHERE score is NULL;

```

# Constraints
```{r}
#Create table where student column *cannot* be NULL
dbGetQuery(mydb,"CREATE TABLE test2 (
  score INTEGER, 
  student TEXT NOT NULL
  );")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")

dbGetQuery(mydb,"CREATE TABLE test2 (
   score INTEGER DEFAULT 0, 
   student TEXT
   );")

dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")
dbGetQuery(mydb,"INSERT INTO test2 (student) VALUES ('B');")

dbGetQuery(mydb, "SELECT * FROM test2;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")

dbGetQuery(mydb,"CREATE TABLE test2 (
  score INTEGER UNIQUE, 
  student TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#Error because of unique
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#NULL is exempt
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")

#EXERCISE 5
#Recreate toy dataset 2 using SQL commands and the constraint that for one of the integer variablesthe default value will be zero.
dbGetQuery(mydb,"CREATE TABLE elameanscore_test2 (
  borough TEXT, 
  school TEXT,
  score INTEGER DEFAULT 0
  );")

dbListTables(mydb)

#Test table by inserting some empty values. 
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West' );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square' );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square' );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School - Washington Heights',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Bronx', 'Success Academy Charter School-Bronx 2',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School - Bensonhurst',624 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Harlem 5',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Bronx', 'Success Academy Charter School-Bronx 2',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Cobble Hill',623 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Harlem 3',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Bronx', 'South Bronx Classical Charter School III',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Crown Heights',622 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Bronx', 'Family Life Academy Charter School II',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Bronx', 'South Bronx Classical Charter School III',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School - Bed Stuy 1',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Bed Stuy 2',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Williamsburg',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Bedford Stuyvesant New Beginnings Charter School',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Prospect Heights',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Prospect Heights', 621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Brooklyn', 'Success Academy Charter School-Flatbush',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Queens', 'Success Academy Charter School - Rosedale',621 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Success Academy Charter School-Hells Kitchen',620 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test2 (borough, school, score) VALUES ( 'Manhattan', 'Sisulu-Walker Charter School of Harlem',620 );")

#Display your new tables. 

dbGetQuery(mydb, "SELECT * FROM elameanscore_test2;")

#Delete your table.  

dbGetQuery(mydb, "DROP TABLE elameanscore_test2;")
dbListTables(mydb)

```


# Adding a column with a default value
```{r}
#Add a column with default value 1
dbGetQuery(mydb, "ALTER TABLE studentAssessment ADD email INTEGER DEFAULT 1 ")
dbGetQuery(mydb, "SELECT * FROM studentAssessment LIMIT 10;")

#Delete a column
dbGetQuery(mydb, "ALTER TABLE studentAssessment DROP COLUMN email;")

#EXERCISE 6
#Add a column to toy data table 2 with a default value of 3. 
dbGetQuery(mydb, "ALTER TABLE elameanscore ADD grade INTEGER DEFAULT 3 ")

#Display your new table. 
dbGetQuery(mydb, "SELECT * FROM elameanscore LIMIT 20;")

#Delete this column.
dbGetQuery(mydb, "ALTER TABLE elameanscore DROP COLUMN grade;")
dbGetQuery(mydb, "SELECT * FROM elameanscore LIMIT 20;")

```


# ID Columns
```{r}
dbGetQuery(mydb,"CREATE TABLE test3 (
  id INTEGER AUTO_INCREMENT PRIMARY KEY, #Not standard syntax
  score INTEGER, 
  student TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (1, 'A');")
dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (5, 'B');")

dbGetQuery(mydb, "SELECT * FROM test3;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test3;")

#EXERCISE 7
#Create a new table with four variables and a primary key that is a sequential id value.
dbGetQuery(mydb,"CREATE TABLE elameanscore_test4 (
  borough TEXT, 
  school TEXT,
  score INTEGER,
  id INTEGER AUTO_INCREMENT PRIMARY KEY
  );")

dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',626,1 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',625,2 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',624,3 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',624,4 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School - Washington Heights',624,5 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Bronx', 'Success Academy Charter School-Bronx 2',624,6 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School - Bensonhurst',624,7 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',623,8 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Harlem 5',623,9 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Bronx', 'Success Academy Charter School-Bronx 2',623,10 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Cobble Hill',623,11 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',622,12 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Harlem 3',622,13 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Bronx', 'South Bronx Classical Charter School III',622,14 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Crown Heights',622,15 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',621,16 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Union Square',621,17 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Upper West',621,18 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Bronx', 'Family Life Academy Charter School II',621,19 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Bronx', 'South Bronx Classical Charter School III',621,20 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School - Bed Stuy 1',621,21 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Bed Stuy 2',621,22 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Williamsburg',621,23 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Bedford Stuyvesant New Beginnings Charter School',621,24 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Prospect Heights',621,25 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Prospect Heights',621,26 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Brooklyn', 'Success Academy Charter School-Flatbush',621,27 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Queens', 'Success Academy Charter School - Rosedale',621,28 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Success Academy Charter School-Hells Kitchen',620,29 );")
dbGetQuery(mydb, "INSERT INTO elameanscore_test4 (borough, school, score, id) VALUES ( 'Manhattan', 'Sisulu-Walker Charter School of Harlem',620,30 );")

dbGetQuery(mydb, "SELECT * FROM elameanscore_test4;")

```

## Filtering (WHERE)
```{r}
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 ORDER BY date_submitted DESC;")

#OR Statement
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 OR date_submitted < 2 ORDER BY date_submitted DESC;")

#AND Statement
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 AND id_student = 325750 ORDER BY date_submitted DESC;")

#LIKE
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region%';")

#Begin with 'Region'
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE 'Region%';")

#End with 'Region'
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region';")

#'c' is the second letter
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '_c%';")

#IN
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region IN ('Wales','Ireland');")

#EXERCISE 8
#Query toy data table 2 to filter for schools that score greater than 622 and are based in Brooklyn.

dbGetQuery(mydb, "SELECT school, borough, score FROM elameanscore WHERE score > 622 AND borough = 'Brooklyn' ORDER BY score DESC;")

dbGetQuery(mydb, "SELECT school, borough, score FROM elameanscore WHERE borough IN ('Queens','Bronx');")


```

## Removing Duplicates
```{r}
dbGetQuery(mydb, "SELECT DISTINCT region FROM studentInfo;")

dbGetQuery(mydb, "SELECT DISTINCT region, gender FROM studentInfo;")

#EXERCISE 9
#Insert a duplicate row into one toy data table 2. 
dbGetQuery(mydb, "INSERT INTO elameanscore VALUES (30, 'Manhattan', 'Sisulu-Walker Charter School Of Harlem', 620 );")
dbGetQuery(mydb, "SELECT * FROM elameanscore;")

#Query the table without including duplicates.
dbGetQuery(mydb, "SELECT DISTINCT school FROM elameanscore;")

```

## Conditional Expressions (non-standard)
```{r}
dbGetQuery(mydb, "CREATE TABLE booltest (a INTEGER, b INTEGER);")
dbGetQuery(mydb, "INSERT INTO booltest VALUES (1, 0);")
dbGetQuery(mydb, "SELECT * FROM booltest;")

dbGetQuery(mydb,"SELECT
  CASE WHEN a THEN 'true' ELSE 'false' END as boolA,
  CASE WHEN b THEN 'true' ELSE 'false' END as boolB
  FROM booltest")

dbGetQuery(mydb,"SELECT
  CASE a WHEN 1 THEN 'true' ELSE 'false' END as boolA,
  CASE b WHEN 1 THEN 'true' ELSE 'false' END as boolB
  FROM booltest")
```

#Relationships (JOIN) - *Slide*
```{r}

#Create two tables with matches and join them

dbGetQuery(mydb, "CREATE TABLE left_table (id INTEGER, description TEXT);")
dbGetQuery(mydb, "CREATE TABLE right_table (id INTEGER, description TEXT);")

dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 1, 'left 01');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 2, 'left 02');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 3, 'left 03');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 4, 'left 04');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 5, 'left 05');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 6, 'left 06');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 7, 'left 07');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 8, 'left 08');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 9, 'left 09');")

dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 6, 'left 06');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 7, 'left 07');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 8, 'left 08');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 9, 'left 09');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 10, 'left 10');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 11, 'left 11');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 12, 'left 12');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 13, 'left 13');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 14, 'left 14');")

dbGetQuery(mydb, "SELECT * FROM left_table;")
dbGetQuery(mydb, "SELECT * FROM right_table;")

dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           JOIN right_table AS r ON l.id = r.id")

dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           RIGHT JOIN right_table AS r ON l.id = r.id")

dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           LEFT JOIN right_table AS r ON l.id = r.id")

#Union
dbGetQuery(mydb, "SELECT * FROM left_table
           UNION
           SELECT * FROM right_table;")


#EXERCISE 10
# Join two toy data tables using common school variable so that your query returns all the values from one table and only those that match from the other. 

dbGetQuery(mydb,"SELECT l.score AS elameanscore, r.score AS mathmeanscore 
           FROM elameanscore AS l 
           JOIN mathmeanscore AS r ON l.school = r.school")

```
```{r}
#Now disconnect from your database
dbDisconnect(mydb)

#Then retunr to your AWS console and:

#1. Click on "Actions" and then "Stop"
#2. Do NOT make a snapshot
#3 Click on "Actions" again and click "Delete"
#4. Unclick "Make a final snapshot"
#5. Clicl "I acknowledge that upon instance deletion, automated backups, including system snapshots and point-in-time recovery, will no longer be available."
#6. Type "delete me" into the field

#Failure to follow these steps could result in charges to your credit card.


```

